Competition for Access to Opportunities between Modes: Multimodal Spatial Availability

### 2023-07-31 TRB-manuscript/TRB-manuscript.tex
The submitted TRB draft

% Repository:  https://github.com/chiehrosswang/TRB_LaTeX_tex
%
% Transportation Research Board conference paper template
% version 4.0 Lite (updates made to be compatible in Overleaf and ShareLaTeX)
%
%
% When numbered option is activated, lines are numbered.
\documentclass[numbered]{trbunofficial}
\usepackage{graphicx}
\usepackage{booktabs}

@@

% Put here what will go to headers as author
\AuthorHeaders{Multimodal spatial availability - the case of low
emission zones in Madrid}
\title{Multimodal spatial availability - the case of low emission zones
in Madrid}

% TODO: add macros for easier formatting of \author.
\author{%
    \textbf{Anastasia Soukhov}\\\textit{Corresponding Author}\\
  PhD Candidate\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:soukhoa@mcmaster.ca}{\nolinkurl{soukhoa@mcmaster.ca}}\\
  \hfill\break
    \textbf{Javier Tarriño Ortiz}\\
  PhD Graduate\\
  Universidad Politécnica de Madrid, Spain\\
  \href{mailto:bob@example.com}{\nolinkurl{bob@example.com}}\\
  \hfill\break
    \textbf{Julio A. Soria-Lara}\\
  Associate Professor\\
  Urban and Regional Planning, Universidad Politécnica de Madrid,
Spain\\
  \href{mailto:derek@example.com}{\nolinkurl{derek@example.com}}\\
  \hfill\break
    \textbf{Antonio Páez}\\
  Associate Professor\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}\\
  \hfill\break
  }

@@

% If words are counted manually, put that number here. This does not include
% figures and tables. This can also be used to avoid problems with texcount
% program i.e. if one does not have it installed.
\TotalWords{7500}

@@

\section{Abstract}
An increasing number of studies within the domain of transportation
planning are concerned with the inequities in accessibility to
opportunities. A dimension of these inequities arises from differences
in access by mode type (e.g., commuting using a car as opposed to
transit). However, methods implemented in current accessibility
literature are lacking within the context of multi-modal analysis. This
paper presents an extension of spatial availability, a
singly-constrained competitive accessibility measure, for the context of
multi-modal accessibility analysis. We first illustrate the features of
spatial availability that lend itself to multi-modal analysis. We then
demonstrate its use on the case study of Low Emission Zones in Madrid
(Spain) and highlight how this policy intervention changes the
accessibility of populations using different modes. In summary, spatial
availability can be used to create and interpret multi-modal policy
intervention scenarios unlike previous methods: this creation and
interpretation can help regions envision a more sustainable and
equitable access-to-opportunity landscape.
\hfill\break%
\hfill\break%
\noindent\textit{Keywords}:  Multimodal, Accessibility, Equity, Policy
scenarios, Low emission zones,  
\newpage

@@

@@

\ascline{1.5pt}{666666}{1-7}\endfirsthead

@@

Implementing urban policies that re-shape cities through accessibility
gains (i.e., the \emph{potential to interact} with opportunities as a
result of land-use mix and transport systems as originally defined by
\citet{hansenHowAccessibilityShapes1959}) have been widely applied
within the transportation literature and is increasingly discussed by
planners
\citep{levinsonTransportAccessManual2020, gowerPlanningInnovationCity2022, siddiqToolsTradeAssessing2021, yanAccessibilityBasedPlanningAddressing2021}.
An important challenge in the identification of interventions that
sustainably and equitably transform cities is the effective evaluation
of \emph{trade-offs}: cities are complex and dynamic ecologies, and
advantaging one component of the city can disadvantage another area,
population, or sub-component In this way, policy evaluation should take
a \emph{systems} approach
\citep{fikselSustainabilityResilienceSystems2006}. One way of
considering systems is from the perspective of the \emph{finite}. As an
illustration, consider the amount of transport space within a city: the
amount is typically finite so re-allocating road space away from one
mode directly impacts the performance of the others (see the literature
on road space reallocation e.g.,
\citet{valencaMainChallengesOpportunities2021}). Evaluating policy
impacts in the context of \emph{finity} provides a way to contextualize
the balance of trade-offs that the citizens of a city should tolerate.

From the perspective of urban transport systems, location-based
accessibility measures have been used in the context of policy
evaluation. For instance, \citet{leeMeasuringImpactsNew2018} assesses
the transit accessibility gains to healthcare and employment
opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA
after the transit system's re-design and introduction of a rapid bus
system. However, a limitation of this study, like others that implement
accessibility measures, is they do not calculate results under a
\emph{constrained} framework i.e., one of \emph{finity}. The citizens of
Columbus should experience qualitative accessibility gains - but is it
at the expense of access to opportunities by other modes? As another
example, \citet{mohriClusteringMethodMeasuring2021} implements a
modified cumulative opportunity measure to assess differences between
private vehicle and transit system accessibility to jobs in Melbourne,
but a similar question remains, does the accessibility afforded to the
private vehicle using population come at the expense of accessibility
losses to transit users?

The two studies discussed in the previous paragraph use
\emph{non-competitive} accessibility measures. There is a branch of
location-based accessibility measures that do incorporate the effect of
competition for opportunities by the population in the region. However,
we argue that these existing methods fall short in acknowledging the
\emph{finity} of opportunities. For instance,
\citet{maoMeasuringSpatialAccessibility2013} applies the competitive
measure, the two-step floating catchment approach (2SFCA) for the case
of access to healthcare services in Florida for both a multimodal
network and a single modal network. While the differences in modal
access are discussed, the question of how the advantage in access
afforded by one mode over another impacts access for different mode
users is unanswered.

This question of how much one mode-using population can access at the
expense of another mode-using population is pertinent equity question in
the evaluation of policy scenarios that are multimodal. For instance,
consider the impact of a low emission zone (LEZ). LEZ is a policy of
spatial and modal discrimination: the circulation of vehicles that are
excessively polluting are restricted in specific areas in a city. In the
recognition that opportunities are finite, the implementation of a LEZ
explicitly reduces the access that the population using polluting
vehicles has to opportunities. This restriction allows the population
using other more sustainable modes to potentially have a higher level of
access than before the LEZ implementation. This evaluation is especially
urgent as LEZ are currently in effect in cities globally; their
reception has been mixed \citep{tarrinoortizPublicAcceptabilityLow2021}
and may be having negative impacts on disadvantaged populations who have
become mobility restricted
\citep{devrijNooneVisitsMe2022, verbeekJustManagementUrban2022}.
Measures that evaluate the accessibility of modes given both
\emph{constrained} and \emph{competitive} considerations are lacking in
the literature, but are needed, to evaluate such policy interventions to
effectively comment on how accessibility changes as a result of the
mobility-restricted mode.

In \citet{soukhovIntroducingSpatialAvailability2023}, we introduce
spatial availability, a type of location-based accessibility measure
that is both \emph{constrained} and \emph{competitive}. In this paper,
we extend the spatial availability measure into a multimodal framework
and explore its use in answering the question outlined: \emph{``given
opportunities are finite, how many are available to a given location
depending on the mode used?''.} The answer to this question quantifies
how many opportunities can be accessed, considering competition, for
different modes. To foreground this exploration, in Section 2, we
discuss short falls of a few existing location-based measures in
comparison to spatial availability through a synthetic example. In
Section 3, the spatial availability of an empirical example of the LEZ
in the city of Madrid, Spain is calculated. We demonstrate how the
restriction of private vehicles with the LEZ implementation impacts the
spatial availability of opportunities for each sub-population using
transit, cycling and walking modes. In Section 4, we provide concluding
remarks on the strengths of the use of spatial availability as a
multi-modal accessibility measure, limitations, and potential future
uses in policy planning scenarios.

\hypertarget{a-brief-review-of-multimodal-accessibility-measurement-methods}{%
\section{A brief review of multimodal accessibility measurement
methods}\label{a-brief-review-of-multimodal-accessibility-measurement-methods}}

The Hansen-type measure does not consider competition between modes nor
is it constrained. As an example, the work of
\citet{tahmasbiMultimodalAccessibilitybasedEquity2019} uses the
Hansen-type measure to measure the potential interaction with retail
locations using walking, public transit, and car modes \(m\). \(S_i^m\)
is the sum of retail locations \(j\) that can potentially be interacted
with under the travel impedance as calculated for each \(i\) and \(m\).
In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
this work, they demonstrate that the car mode has the highest
\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
a car can potentially interact with the most retail opportunities than
populations using other modes. However, the higher \(S_i^{m=car}\)
values are not a result of lower \(S_i^{m}\) values for other modes: it
is not assumed that car-using populations potentially accessing more
opportunities take away potential opportunities for other populations
within the measure. Put another way, this measure does not consider
competition. This measure is also not constrained: there is no global
maximum for \(S_i\) or \(S_i^m\) values, they are presented as a
population normalized accessibility index. This makes the interpretation
of the `potentially interacted opportunities' relative to the region,
making comparisons of the results across different regions challenging.

However, opportunities in a region can be considered finite. There are
only so many school-seats, hospital capacity, food stores, jobs, etc.,
in a region and if one person interacts with an opportunity at a given
time, it is taken. As such, if one person is advantaged and has the
ability to reach more opportunities through a lower travel-cost mode,
than they have more opportunities to potentially interact with more
opportunities than other people. From the other perspective, their are
fewer opportunities left to be potentially interacted with for
populations using higher travel-cost modes. In this way, populations
using modes with a higher travel impedance are at a higher access
disadvantage than populations using lower travel impedance modes. This
recognition is the motivation behind integrating \emph{competition} for
opportunities within multimodal accessibility measures. Arguably one of
the most popular competitive location-based accessibility measures is
the two-step floating catchment area (2SFCA) approach popularized by
\citet{luoMeasuresSpatialAccessibility2003} who simplified the approach
proposed by \citet{shenLocationCharacteristicsInnercity1998} (with
similar considerations for competition in
\citet{weibullAxiomaticApproachMeasurement1976} and
\citet{josephMeasuringPotentialPhysical1982}).

To illustrate, \citet{taoInvestigatingImpactsPublic2020a} calculates
\(a_i^m\) to jobs for different income-group populations in Shenzhen
(China) using \(m = \text{public transit}\) and \(m={car}\). They
demonstrate that \(i\)s with low-income populations have lower \(a_i^m\)
than \(i\)s with higher-income populations. Further, they demonstrate
that \(a_i^{m=\text{public transit}}\) is lower than
\(a_i^{m=\text{car}}\) at many \(i\)s, arguing that this may put \(i\)s
with lower-income populations in a further disadvantage. \(a_i\) and/or
\(a_i^m\) are used to compare relative spatial differences in overall
competitive accessibility and modal competitive accessibility, but
because there is no global maximum, making it is difficult to interpret
the significance between differences in \(a_i^{m}\) values. Questions
such as: What is the impact that competition has on the difference in
\(a_i^m\) values? How does impact vary spatially? And what is the
interpretation of this difference? are left unanswered.

>>>>

% Repository:  https://github.com/chiehrosswang/TRB_LaTeX_tex
%
% Transportation Research Board conference paper template
% version 4.0 Lite (updates made to be compatible in Overleaf and ShareLaTeX)
%
%
% When numbered option is activated, lines are numbered.
\documentclass[]{trbunofficial}
\usepackage{graphicx}
\usepackage{booktabs}

@@

% Put here what will go to headers as author
\AuthorHeaders{}
\title{Competition for Access to Opportunities between Modes: Multimodal
Spatial Availability}

% TODO: add macros for easier formatting of \author.
\author{%
    \textbf{Anastasia Soukhov}\\\textit{Corresponding Author}\\
  PhD Candidate\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:soukhoa@mcmaster.ca}{\nolinkurl{soukhoa@mcmaster.ca}}\\
  \hfill\break
    \textbf{Javier Tarriño-Ortiz}\\
  PhD Graduate\\
  Universidad Politécnica de Madrid, Spain\\
  \href{mailto:javier.tarrino.ortiz@upm.es}{\nolinkurl{javier.tarrino.ortiz@upm.es}}\\
  \hfill\break
    \textbf{Julio A. Soria-Lara}\\
  Associate Professor\\
  Urban and Regional Planning, Universidad Politécnica de Madrid,
Spain\\
  \href{mailto:julio.soria-lara@upm.es}{\nolinkurl{julio.soria-lara@upm.es}}\\
  \hfill\break
    \textbf{Antonio Páez}\\
  Associate Professor\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}\\
  \hfill\break
  }

@@

% If words are counted manually, put that number here. This does not include
% figures and tables. This can also be used to avoid problems with texcount
% program i.e. if one does not have it installed.
\TotalWords{7195}

@@

\section{Abstract}
An increasing number of studies within the domain of transportation
planning are concerned with the inequities in accessibility to
opportunities. A dimension of these inequities arises from differences
in access by mode type (e.g., commuting using a car as opposed to
transit). However, methods implemented in current accessibility
literature are lacking within the context of multimodal analysis. This
paper presents an extension of spatial availability, a
singly-constrained competitive accessibility measure, for the context of
multimodal accessibility analysis. We first illustrate the features of
spatial availability that lend itself to multimodal analysis. We then
demonstrate its use on the case study of Low Emission Zones in Madrid
(Spain) and highlight how this policy intervention changes the
accessibility of populations using different modes. In summary, spatial
availability can be used to create and interpret multimodal policy
intervention scenarios unlike previous methods: this creation and
interpretation can help regions envision a more sustainable and
equitable access-to-opportunity landscape.
\hfill\break%
\hfill\break%
\noindent\textit{Keywords}:  Multimodal, Accessibility, Equity, Policy
scenarios, Low emission zones,  
\newpage

Implementing urban policies that re-shape cities through accessibility
gains (i.e., the \emph{potential to interact} with opportunities as a
result of land-use mix and transport systems as originally defined by
\citet{hansenHowAccessibilityShapes1959}) have been widely applied
within the transportation literature and is increasingly discussed by
planners
\citep{levinsonTransportAccessManual2020, gowerPlanningInnovationCity2022, siddiqToolsTradeAssessing2021, yanAccessibilityBasedPlanningAddressing2021}.
An important challenge in the identification of interventions that
equitably transform cities is the effective evaluation of
\emph{trade-offs}: cities are complex and dynamic ecologies, and
advantaging one component of the city can disadvantage another area,
population, or sub-component In this way, policy evaluation should take
a \emph{systems} approach
\citep{fikselSustainabilityResilienceSystems2006}. One way of
considering systems is from the perspective of the \emph{finite}. As an
illustration, consider the amount of transport space within a city: the
amount is typically finite so re-allocating road space away from one
mode directly impacts the performance of the others (see the literature
on road space reallocation e.g.,
\citet{valencaMainChallengesOpportunities2021}). Evaluating policy
impacts in the context of \emph{finity} provides a way to contextualize
the balance of trade-offs that the citizens of a city should tolerate.

From the perspective of urban transport systems, location-based
accessibility measures have been used in the context of policy
evaluation. For instance, \citet{leeMeasuringImpactsNew2018} assesses
the transit accessibility gains to healthcare and employment
opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA
after the transit system's re-design and introduction of a rapid bus
system. However, a limitation of this study, like others that implement
accessibility measures, is they do not calculate results under a
\emph{constrained} framework i.e., one of \emph{finity}. The citizens of
Columbus should experience qualitative accessibility gains - but is it
at the expense of access to opportunities to populations using other
modes?

The work of \citet{leeMeasuringImpactsNew2018} uses
\emph{non-competitive} accessibility measures. There is a branch of
location-based accessibility measures that do incorporate the effect of
competition for opportunities by the population in the region. However,
we argue that these existing methods fall short in acknowledging the
\emph{finity} of opportunities. For instance,
\citet{maoMeasuringSpatialAccessibility2013} applies a competitive
measure, two-step floating catchment approach (2SFCA), for the case of
access to healthcare services in Florida for both a multimodal network
and a single modal network. While the differences in modal access are
discussed, the question of how the advantage in access afforded by one
mode over another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the
expense of another mode-using population is a pertinent equity question
in the evaluation of policy scenarios that are multimodal. For instance,
consider the impact of a low emission zone (LEZ). LEZ is a policy of
spatial and modal discrimination: the circulation of vehicles that are
excessively polluting are restricted in specific regions. In the
recognition that opportunities are finite, the implementation of a LEZ
explicitly reduces the access that the population using polluting
vehicles has to opportunities. This restriction allows the population
using other more sustainable modes to potentially have a higher level of
access than before the LEZ implementation. This evaluation is especially
urgent as LEZ are currently in effect in cities globally; their
reception has been mixed \citep{tarrinoortizPublicAcceptabilityLow2021}
and may be having negative impacts on disadvantaged populations who have
become mobility restricted
\citep{devrijNooneVisitsMe2022, verbeekJustManagementUrban2022}.
Measures that evaluate the accessibility of modes given both
\emph{constrained} and \emph{competitive} considerations are lacking in
the literature, but are needed, to evaluate such policy interventions
impact on accessibility.

In \citet{soukhovIntroducingSpatialAvailability2023}, we introduce
spatial availability, a type of location-based accessibility measure
that is both \emph{constrained} and \emph{competitive}. In this paper,
we extend the spatial availability measure into a multimodal framework
and explore its use in answering the question outlined: \emph{``given
opportunities are finite, how many are available to a given location
depending on the mode used?''.} The answer to this question quantifies
how many opportunities can be accessed, considering competition, for
different modes. To foreground this exploration, in Section 2, we
discuss short falls of a few existing location-based measures in
comparison to spatial availability through a synthetic example. In
Section 3, the spatial availability of an empirical example of the LEZ
in the city of Madrid, Spain is calculated. We demonstrate how the
restriction of car circulation could have impacted the spatial
availability of opportunities for each sub-population using transit,
cycling and walking modes. In Section 4, we provide concluding remarks
on the strengths of the use of spatial availability as a multimodal
accessibility measure and potential future uses in policy planning
scenarios.

\hypertarget{a-review-of-multimodal-accessibility-measures}{%
\section{A review of multimodal accessibility
measures}\label{a-review-of-multimodal-accessibility-measures}}

The Hansen-type measure does not consider competition between modes nor
is it constrained. As an example, the work of
\citet{tahmasbiMultimodalAccessibilitybasedEquity2019} uses the
Hansen-type measure to measure the potential interaction with retail
locations using walking, public transit, and car modes \(m\). \(S_i^m\)
is the sum of retail locations \(j\) that can potentially be interacted
with under the travel impedance as calculated for each \(i\) and \(m\).
In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
this work, they demonstrate that the car mode has the highest
\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
a car can potentially interact with the most retail opportunities than
populations using other modes. However, the higher \(S_i^{m=car}\)
values are not a result of lower \(S_i^{m}\) values for other modes: it
is not assumed that car-using populations potentially accessing more
opportunities take away potential opportunities for other populations
within the measure (no consideration for competition). This measure is
also not constrained: there is no global maximum for \(S_i\) or
\(S_i^m\) values, they are presented as a population normalized
accessibility index. This makes the interpretation of the `potentially
interacted opportunities' relative to the region, making comparisons of
the results across different regions challenging.

However, opportunities in a region can be considered finite. There are
only so many school-seats, hospital capacity, jobs, etc., in a region
and if one person interacts with an opportunity at a given time, it is
taken. As such, if one person is advantaged and has the ability to reach
more opportunities through a lower travel-cost mode, than they have more
opportunities to potentially interact with more opportunities than other
people. From the other perspective, their are fewer opportunities left
to be potentially interacted with for populations using higher
travel-cost modes. In this way, populations using modes with a higher
travel impedance are at a higher access disadvantage than populations
using lower travel impedance modes. This recognition is the motivation
behind integrating \emph{competition} for opportunities within
multimodal accessibility measures. Arguably one of the most popular
competitive location-based accessibility measures is the two-step
floating catchment area (2SFCA) approach popularized by
\citet{luoMeasuresSpatialAccessibility2003} who simplified the approach
proposed by \citet{shenLocationCharacteristicsInnercity1998} (with
similar considerations for competition in
\citet{weibullAxiomaticApproachMeasurement1976} and
\citet{josephMeasuringPotentialPhysical1982}).

To illustrate, \citet{taoInvestigatingImpactsPublic2020a} calculates
\(a_i^m\) to jobs for different income-group populations in Shenzhen
(China) using \(m = \text{public transit}\) and \(m={car}\). They
demonstrate that \(i\) with low-income populations have lower \(a_i^m\)
than \(i\) with higher-income populations. Further, they demonstrate
that \(a_i^{m=\text{public transit}}\) is lower than
\(a_i^{m=\text{car}}\) at many \(i\), arguing that this may put \(i\)
with lower-income populations in a further disadvantage. \(a_i\) and/or
\(a_i^m\) are used to compare relative spatial differences in overall
competitive accessibility and modal competitive accessibility, but
because there is no global maximum, making it is difficult to interpret
the significance between differences in \(a_i^{m}\) values. Questions
such as: what is the impact that competition has on the difference in
\(a_i^m\) values? How does impact vary spatially? And what is the
interpretation of this difference? are left unanswered.

Spatial availability improves on previous multimodal accessibility
approaches as it considers \emph{competition} in the potential
interaction with opportunities in a \emph{constrained} framework (e.g.,
finite opportunities). This is done by considering: 1) competition
between mass effect (e.g., the advantage of sub-populations residing in
relatively low population-density and high opportunity-proximate areas)
and 2) competition between travel impedance (e.g., sub-populations with
relatively low travel-impedance) through a proportional allocation
mechanism. The following sub-section demonstrates how spatial
availability compares to the Hansen-type and Shen-type measures through
a synthetic example.

\hypertarget{multimodal-spatial-availability-v_im}{%
\subsection{\texorpdfstring{Multimodal spatial availability
\(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}

In brief, we define the \emph{spatial availability} at \(i\) ( \(V_{i}\)
) as the proportion of all opportunities in the region \(O\) that are
allocated to location \(i\) from all opportunity destinations \(j\).
\(V_{i}\) is a value of how many opportunities are available to each
location \(i\) out of all the opportunities in the region. The general
formulation of spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}):

Spatial availability measure is introduced in
\citet{soukhovIntroducingSpatialAvailability2023}. The unique feature in
the measure is the balancing factor \(F^t_{ij}\), a proportional
allocation mechanisms, that ensures \(V_i\) calculated for each \(i\)
sums to the total number of opportunities. Through \(F^t_{ij}\), spatial
availability is a \emph{competitive} and \emph{constrained}
accessibility measure that handles the number of opportunities in the
region in a finite way. \(F^t_{ij}\) consists of two components: a
population-based balancing factor \(F^p_{i} = \frac{P_i}{\sum_i P_i}\)
and an impedance-based balancing factor
\(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that, respectively,
allocate opportunities to \(i\) in proportion to the size of the
population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect).

\(F^p_{i}\) and \$ F\^{}c\_\{ij\}\$ are calculated for each \(i\) such
that they both equal 1 when summed across all \(i\) in the region (e.g.,
\(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
factors are combined multiplicatively to yield \(F^t_{ij}\) which
ensures that a proportion of the opportunities \(O_j\) are allocated to
each \(i\) accordingly. In other words, assuming a finite number of
opportunities in the region, \(F^t_{ij}\) proportionally allocates
\(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
the number of opportunities \emph{spatially available} to the population
at \(i\). This value can be seen to represent spatial availability as it
is a proportion of the opportunities in the region (i.e.,
\(\sum_j O_j = \sum_i V_i\)).

\begin{itemize}
\tightlist
\item
  The population balancing factor for each \(m\) at each \(i\) is
  \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
\item
  The cost of travel balancing factor for each \(m\) at \(i\) is
  \(F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}\)
\end{itemize}

\begin{itemize}
\tightlist
\item
  \(m\) is a set of modes used by populations in the region.
\item
  \(F^{tm}_{ij}\) is a balancing factor \(F^t_{ij}\) for each \(m\) at
  each \(i\).
\item
  \(V^m_{i}\) is the spatial availability \(V_{i}\) for mode \(m\) at
  each \(i\); the sum of \(V^m_{i}\) for all \(m\) at each \(i\) is
  equivalent to the total sum of opportunities in the region (i.e.,
  \(\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}\))
\end{itemize}

\hypertarget{synthetic-example}{%
\subsection{Synthetic example}\label{synthetic-example}}

Consider the following: Figure \ref{fig:Fig1} depicts a region with
population and jobs at three population centers (\(A\), \(B\), \(C\))
and three employment centers (\(1\), \(2\), \(3\)). The population at
each population center is divided into two sub-populations, one using a
faster mode \(z\) and another using a slower mode \(x\), to travel to
employment centers. Population center \(A\) is Suburban: it is closest
to its own relatively large employment center at \(1\), close to the
Urban's equally large employment center \(2\), and has a population that
is smaller than the Urban \(B\) and larger than the Satellite \(C\).
\(B\) has the largest \(x\)-using population, followed by then \(A\),
then \(C\). This synthetic example was inspired by the single-mode
example used in \citet{shenLocationCharacteristicsInnercity1998} and
reconfigured in \citet{soukhovIntroducingSpatialAvailability2023}.

\caption{\label{fig:Fig1} Multimodal synthetic example: locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower mode x and faster mode z).}\label{fig:synthetic-example-plot}
\end{figure}

From the perspective of access to a \emph{finite} amount of
opportunities in the region (\(210,000\) jobs), the sub-population that
is most proximate to jobs, furthest from densely populated centers, and
is using the lowest travel-cost mode \(z\) can potentially access the
most job opportunities. This appears to be the sub-population at \(A\)
using \(z\). From the other perspective, sub-populations located in
opposite conditions (i.e., further away from jobs, close to dense
populations, and using \(x\)) are at a relative job opportunity access
\emph{disadvantage}. From the perspective of inequities, the competition
for opportunities between different mode-using populations matters as it
reflects how well the land-use and transport system serves (or doesn't
serve) them.

@@

\caption{Accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i\ for\ the\ synthetic\ example.}\\

@@

\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i\ for\ the\ synthetic\ example.}\\

@@

The calculated \(S_i^m\), \(a_i^m\) and \(V_i^m\) accessibility values
for each \(i\) and \(m\) are shown in the middle three columns and are
aggregated for each \(i\) in the final two columns in Table 1 . We use a
negative exponential impedance function
\(f(c_{ij}) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.

The Hansen-type measure \(S_i^m\) is presented for each origin and mode
in third column of Table 1 . For all \(i\), the \(z\)-using
sub-population has higher \(S_i^m\) values than the \(x\)-using
sub-populations. Additionally, \(S_i^m\) is equal for both mode-using
populations in \(A\) and \(B\). This is the case because \(S_i^m\) does
not consider \emph{competition}, it only relies on reflecting the count
of opportunities that may be interacted with as a product of
\(f^m(c_{ij}^m)\). Recall, populations in \(A\) and \(B\) have the same
travel impedance to employment centers \(1\), \(2\) and \(3\) (either
15, 30, or 100 minutes using \(x\) or 10, 25, or 80 minutes using
\(z\)). As such, these the calculated \(S_i^m\) values are the same for
both \(A\) and \(B\). Furthermore, the total sum of \(S_i^m\) in the
region is equal to 150,570.2. This value is difficult to interpret: it
represents the weighted sum of opportunities that may be interacted with
within the region based on travel impedance. It cannot be interpreted as
any sort of benchmark since the measure is \emph{non-constrained}. To
connect this example to literature, \(S_i^m\) is calculated in the work
of \citet{tahmasbiMultimodalAccessibilitybasedEquity2019}; they compare
differences in \(S_i^m\) values between modes in a relative and
comparative sense, but make no further interpretation of the \(S_i^m\)
values.

In the fourth and sixth column in Table 1 the Shen-type measure is
calculated: first for both origin and mode \(a_i^m\) as well as
aggregated by the weighted mean mode-population (
\(\sum_m \frac{P_i^m}{P_i}*a_i^m\) ) to represent a value for each
origin \(a_i\). Unlike \(S_i^m\), this measure considers
\emph{competition}. For instance, the \(x\)-using populations in \(A\)
and \(B\) centers do not have the same \(a_i^m\) values as the
\(z\)-using. In fact, \(A\) has the highest values \(a_i^m\) and \(a_i\)
values since this center has the smallest travel impedance to
opportunities (lower than at \(C\), \(A\) and \(B\) are equal) and has
one of these lowest proximity to a relatively high amount of population
(lower than at \(B\)).

However, the Shen-type measure is \emph{non-constrained}: the total sum
of \(a_i^m\) or \(a_i\) is practically meaningless since it represents a
sum of ratios. For instance, the \(z\)-using sub-population at \(A\) has
a value of 1.56 potential jobs per potential job-seeking population
compared to 0.95 for \(x\)-using sub-population. What is the
significance of these values? The difference between these modes is
equal to 0.62, but 0.62 of what? How many more job opportunities are
\(z\) users interacting with than \(x\) users? When \(a_i^m\) is
aggregated to \(a_i\) as shown in the sixth column, the values face
similar interpretability issues.

The Shen-type measure is implemented in the previously discussed work of
\citet{taoInvestigatingImpactsPublic2020a} to calculate modal \(a_i^m\)
values and the aggregated \(a_i\) is implemented in the work of
\citet{carpentieriMultimodalAccessibilityPrimary2020}. However, similar
to the Hansen-type measure, these works discuss relative and spatially
comparative differences in values, they do not make further
interpretation of the \(a_i^m\) or \(a_i\) themselves. This may be
because the Shen-type measure is \emph{non-constrained}, this is no
benchmark or global maximum to which comparisons can be drawn from.

By contrast, spatial availability \(V_i\) considers competition and is
constrained such that the total sum of values is equal to the total
number of opportunities in the region (i.e., \(210,000\) jobs). Seen in
fifth column of Table 1 , \(V_i^m\) for the same mode-using populations
in \(A\) and \(B\) are not the same (as this measure considers
competition). In fact, at \(A\), the \(z\)-using sub-population captures
27,953.18 more spatially available jobs (of the \(210,000\) jobs in the
region) than the sub-population using mode \(x\). The numerical
difference has a practical interpretation.

Furthermore, \(V_i^m\) values for an \(i\) can be aggregated across
\(m\) and compared across \(i\) ( \(V_i = \sum_m{\sum_i{V_i^m}}\) ) as a
result of the proportional allocation mechanism. This aggregation,
\(V_i\), is shown in the seventh column in Table 1 . Again looking at
center \(A\), \(A\) is allocated 65,872.91 spatially available
opportunities for both modes. 71\% of this spatial availability
allocated to \(A\) is assigned to the \(z\)-using population despite
representing 66\% of \(A\)'s population.

Spatial availability can be further aggregated to better interpret
competition between modes. Across the entire region, 137,500 people use
\(z\) (65\% of the region population). However, the \(z\)-using
population accounts for 75\% of the region's total spatial availability
- the rest is allocated to the \(x\)-using population (35\%of the total
population). Notably, the \(x\)-using population captures 10\% less
spatial availability to opportunities than its population proportion.
This understanding can lead us to ask normative questions such as, how
unequal should opportunity access for the two mode-using populations be?
Can the lower-travel-cost populations spare some spatial availability if
a policy of modal-restriction (like a LEZ) was introduced?

Alternatively, inequities in spatial availability between mode-using
populations can be explored through proportional benchmarks. A spatial
availability per capita \(v_i^m\) as presented in Equation
(\ref{eq:SA-per-capita}):

\begin{equation}
\label{eq:SA-per-capita}
v_{i}^m = \frac{V_{i}^m}{P_{i}^m}
\end{equation}

The \(v_i^m\) values for \(A\), \(B\), and \(C\) for the \(x\)-using
sub-populations are 0.95, 0.62 and 0.68 spatially available jobs per
capita, respectively. The \(v_i^m\) for the \(z\)-using sub-populations
are much higher, with values of: 1.56, 1.03 and 1.12 respectively. The
\(x\)-using population, especially at \(B\) and \(C\), are directly
impacted by the jobs that are spatially available to the \(z\)-using
population \emph{in addition to} the mass effect (occurring at \(B\),
high population density) and high travel impedance (occurring at the
Satellite \(C\)).

If, lets say, the planning goal is to have one spatially available job
per mode-using population, a policy intervention can be put in place, to
reduce the \(v_i^z\) values and increase \(v_i^x\) values. This
demonstration is to show how simply the \(V_i^m\) framework can be
manipulated quantify the competitive (dis)advantage in a multimodal
application. In what follows, we further explore competition between
multiple modes through an empirical example.

\hypertarget{empirical-example-madrid-lez}{%
\section{Empirical example: Madrid
LEZ}\label{empirical-example-madrid-lez}}

\hypertarget{multimodal-data-and-methods}{%
\subsection{Multimodal data and
methods}\label{multimodal-data-and-methods}}

Low emission zones (LEZ) have been implemented as a climate change
policy intervention to reduce GHG emissions, improve air quality, and
support sustainable mobility in many countries. Though rules vary, LEZ
aim to deter/reduce traffic in designated zones under threat of penalty
(e.g., fines, seizure of vehicle). From the perspective of restriction
for passenger transport, LEZ are a policy of \emph{geographic
discrimination}. LEZ actively change how people access opportunities by
making the travel impedance more costly for car-mode users. If seeing
opportunities as finite, this discrimination allows populations to
access opportunities by other modes more readily than before. In this
way, LEZ change the multimodal competitive accessibility landscape of a
city.

Spain is one of a few countries with active LEZ and plans to expand
their implementation as specified in their climate-change-related plans:
\emph{Plan Nacional Integrado de Energía y Clima 2021-2030}
\citep{espanaPlanNacionalIntegrado2020} and \emph{Plan Nacional de
Control de la Contaminación Atmosférica}
\citep{espanaResolucion10Enero2020}. Specifically, the national Spanish
law 7/2021 ( \emph{Ley de Cambio Climático y Transición Energética})
will require all municipalities to implement LEZ by 2023 if they meet at
least one of the following requirements: (i) municipalities
\textgreater50,000 inhab.; (ii) islands; and (iii) municipalities
\textgreater{} 20,000 inhab. when air quality exceeds limits specified
in \emph{RD 102/2011 de Mejora de Calidad del Aire}
\citep{barcelonaGUIATECNICAPARA2021}.
